---
title: "PDA_Project_US_Car_Accidents_Analysis"
author: "Group13"
date: "2023-11-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# US Car Accident Analysis

### Inlcuded Libraries

```{r echo = FALSE}
library(tidyverse)
library(lubridate)
library(dplyr)
library(ggplot2)
library(tidyr)
require(ISLR)
library(MASS)
library(MLmetrics)
library(corrplot)
```



# PART I: Data Preperation and cleaning


```{r}
# read the file into a variable
USAcaraccidents = read.csv("C:/Users/anush/OneDrive/Documents/Data Analytics/US_Accidents_March23_sampled_500k.csv")
```

```{r}
head(USAcaraccidents)
```

```{r}
#Checking for Null Values
sum(is.na(USAcaraccidents))
```
```{r}
#Removing null value rows
USAcaraccidents <- USAcaraccidents[complete.cases(USAcaraccidents), ]
```

```{r}
#Confirming cleanup
sum(is.na(USAcaraccidents))
```
```{r}
#Total size
nrow(USAcaraccidents)
```

```{r}
summary(USAcaraccidents)
```




# PART II: Exploratory Data Analysis

### 1.Temperature Distribution Analysis:

* The overlay histogram and boxplot reveal that accidents with severity 2 tend to occur more frequently in extreme temperature conditions compared to severity 3 accidents.

```{r}
# Overlay Histogram with Severity
hist(USAcaraccidents$Temperature[USAcaraccidents$Severity == 2], col = rgb(1, 0, 0, 0.5), main = "Overlay Histogram", xlab = "Temperature (F)", ylab = "Frequency")
hist(USAcaraccidents$Temperature[USAcaraccidents$Severity == 3], col = rgb(0, 0, 1, 0.5), add = TRUE)
legend("topright", legend = c("Severity 2", "Severity 3"), fill = c(rgb(1, 0, 0, 0.5), rgb(0, 0, 1, 0.5)))

```

```{r}
# Boxplot for Temperature
boxplot(USAcaraccidents$Temperature, main = "Temperature Distribution", ylab = "Temperature (F)")
```

### 2.Weather Conditions Distribution Bar Plot:

* Fair/Windy weather conditions dominate the dataset, followed by cloudy conditions. Adverse weather conditions are less common, indicating that most accidents occur in relatively clear weather.
```{r}
# Bar plot for Weather Conditions
weather_counts <- table(USAcaraccidents$Weather_Condition)
barplot(weather_counts, main = "Weather Conditions Distribution", ylab = "Count", col = "lightgreen", las = 2)

```
### 3.Hourly Accident Distribution Bar Plot:

* Accidents are relatively evenly distributed across different hours of the day, with a slight increase during the afternoon.

```{r}
# Extracting the hour from Start_Time
USAcaraccidents$Hour <- as.POSIXlt(USAcaraccidents$Start_Time)$hour

# Hourly accident distribution
hourly_counts <- table(USAcaraccidents$Hour)
barplot(hourly_counts, main = "Hourly Accident Distribution", xlab = "Hour", ylab = "Count", col = "lightcoral")

```


### 4.Accidents by Location Scatter Plot:

* The scatter plot shows the geographical distribution of accidents, highlighting potential hot-spots or areas with higher accident frequencies.

```{r}
# Scatter plot of accidents based on latitude and longitude with transparency and smaller points
plot(USAcaraccidents$Start_Lng, USAcaraccidents$Start_Lat, col = rgb(0, 0, 1, 0.1), pch = 16, main = "Accidents by Location", xlab = "Longitude", ylab = "Latitude")

```

### 5.Temperature by Severity Boxplot:

The boxplot suggests that there is a variation in temperature for different severity levels, with severity 3 accidents showing a wider range of temperatures.
```{r}
boxplot(Temperature.F. ~ Severity, data = USAcaraccidents, main = "Temperature by Severity", xlab = "Severity", ylab = "Temperature (F)", col = "lightblue")

```

### 7.Temperature Distribution Histogram:

The histogram provides a clear overview of the temperature distribution in accidents, helping identify common temperature ranges.
```{r}
# Histogram for Temperature
hist(USAcaraccidents$Temperature.F., main = "Distribution of Temperature", xlab = "Temperature (F)", col = "lightblue")

```

### 8.Visibility Category Bar Plot:

Accidents are more frequent in higher visibility conditions, with the majority falling within the 5-10 miles visibility range.
```{r}
# Create Visibility Categories
USAcaraccidents$Visibility_Category <- cut(USAcaraccidents$Visibility.mi., breaks = c(0, 1, 2, 3, 4, 5, 10, Inf), labels = c("0-1", "1-2", "2-3", "3-4", "4-5", "5-10", "10+"))

# Bar Plot for Visibility Categories
barplot(table(USAcaraccidents$Visibility_Category), 
        main = "Accident Frequency by Visibility Category",
        xlab = "Visibility Category",
        ylab = "Number of Accidents",
        col = "skyblue")


```


### 9.Severity Distribution Bar Plot (ggplot2):

Using ggplot2, the bar plot reiterates the distribution of accident severity, providing an alternative visualization.
```{r}
ggplot(data = USAcaraccidents, aes(x = Severity)) +
  geom_bar(fill = "skyblue") +
  ggtitle("Distribution of Severity")

```


### 10.Impact of Traffic Signals on Severity Stacked Bar Plot:

The stacked bar plot indicates that the presence of traffic signals has a marginal impact on accident severity, with severity 2 being slightly higher at locations with traffic signals.
```{r}

# Assuming 'Traffic_Signal' is the column indicating the presence of traffic signals
# Assuming 'Severity' is the column indicating the severity of accidents

# Convert 'Severity' to factor
USAcaraccidents$Severity <- as.factor(USAcaraccidents$Severity)

# Create a new column to categorize locations with and without traffic signals
USAcaraccidents$Signal_Category <- ifelse(USAcaraccidents$Traffic_Signal, "With Traffic Signal", "Without Traffic Signal")

# Create a bar plot to compare severity distribution
ggplot(data = USAcaraccidents, aes(x = Signal_Category, fill = Severity)) +
  geom_bar(position = "fill", show.legend = TRUE) +
  ggtitle("Impact of Traffic Signals on Accident Severity") +
  xlab("Traffic Signal Presence") +
  ylab("Proportion of Accidents") +
  scale_fill_manual(values = c("1" = "red", "2" = "orange", "3" = "yellow", "4" = "green"))



```

### 11.Accident Trends Over Years Time Series Plot:


* The time series plot illustrates the overall trend in accident frequency over the years, revealing any notable patterns or changes.
```{r}


# Assuming 'Start_Time' is the column indicating the start time of accidents
# Convert 'Start_Time' to POSIXct format
USAcaraccidents$Start_Time <- as.POSIXct(USAcaraccidents$Start_Time, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

# Extract year and month for time series analysis
USAcaraccidents$YearMonth <- format(USAcaraccidents$Start_Time, "%Y-%m")

# Time series plot for accident frequency over years
ggplot(data = USAcaraccidents, aes(x = Start_Time)) +
  geom_freqpoly(binwidth = 60*60*24*365, color = "blue") +
  ggtitle("Accident Trends Over Years") +
  xlab("Year") +
  ylab("Accident Frequency")

```



### 12.Accident Frequency by Day of Week Stacked Bar Plot:

* Accidents are fairly evenly distributed throughout the week, with a slight increase during weekdays compared to weekends.
```{r}
ggplot(USAcaraccidents, aes(x = factor(wday(Start_Time, label = TRUE, abbr = FALSE)), fill = factor(Severity))) +
  geom_bar(position = "stack") +
  ggtitle("Accident Frequency by Day of Week") +
  xlab("Day of Week") +
  ylab("Count") +
  scale_fill_manual(values = c("1" = "green", "2" = "yellow", "3" = "red", "4" = "purple"))

```



# PART III: Hypothesis Testing

```{r}
severity <- USAcaraccidents$Severity
```

```{r}
temperature <- USAcaraccidents$Temperature.F.
```

> Null Hypothesis - There is no significant difference in accident severity between tempratures above 50 and temprature 50 or below.
<br>
> Alternate Hypothesis - There is a significant difference in accident severity between tempratures above 50 and temprature 50 or below.

```{r}
severity_above_threshold <- severity[temperature > 50]
severity_below_threshold <- severity[temperature <= 50]
```

```{r}
severity_above_threshold <- as.numeric(severity_above_threshold)
severity_below_threshold <- as.numeric(severity_below_threshold)
t_test_result <- t.test(severity_above_threshold, severity_below_threshold)
t_test_result
```

```{r}
alpha <- 0.05
if (t_test_result$p.value < alpha) {
  print("Reject the null hypothesis. There is a significant difference in accident severity based on temperature.")
} else {
  print("Fail to reject the null hypothesis. No significant difference in accident severity based on temperature.")
}
```

```{r}
day_night <- USAcaraccidents$Sunrise_Sunset
```

> Null Hypothesis (H0) - There is no significant difference in accident severity between day and night.
<br>
Alternate Hypothesis (H1) - There is a significant difference in accident severity between day and night.


```{r}
severity_day <- severity[day_night == "Day"]
severity_night <- severity[day_night == "Night"]
severity_day <- as.numeric(severity_day)
severity_night <- as.numeric(severity_night)
t_test_result <- t.test(severity_day, severity_night)
t_test_result
```

```{r}
alpha <- 0.05
if (t_test_result$p.value < alpha) {
  print("Reject the null hypothesis. There is a significant difference in accident severity between day and night.")
} else {
  print("Fail to reject the null hypothesis. No significant difference in accident severity between day and night.")
}
```




# PART IV: Linear Regression Modals


```{r}
#Creating a variable containing only continuous numerical variables
selected_columns <- c('Distance.mi.', 'Temperature.F.', 'Wind_Chill.F.','Humidity...', 'Pressure.in.','Visibility.mi.', 'Wind_Speed.mph.', 'Precipitation.in.')
```

```{r}
pairs(USAcaraccidents[selected_columns], col ="blue")
```

```{r}
# Calculate the correlation matrix for numeric variables
correlation_matrix <- cor(USAcaraccidents[selected_columns])
# Print the correlation matrix
print(correlation_matrix)
```

>> We can observe that Temperature, Wind Chill, Pressure and Visibility are inversely related to Distance.
<br>
>>Pressure has the highest correlation with Distance but that does not make sense. In the correlation table we observe that pressure has a positive relation with temperature, which doesn't seem possible. So we can mark our observations here and continue with other


### Let's create a simple Linear regression model

```{r}
Numerical_accident_data <- USAcaraccidents[selected_columns]
summary(Numerical_accident_data)
```


```{r}
# Splitting the dataset into 2
train_index = sample(2,nrow(Numerical_accident_data),replace=TRUE, prob = c(0.8,0.2))
Accident_Training <- Numerical_accident_data[train_index==1,]
Accident_Testing <- Numerical_accident_data[train_index==2,]
dim(Accident_Training)
dim(Accident_Testing)
```



```{r}
#Creating a simple linear regression modal
lm_model1 <- lm(Distance.mi. ~ Precipitation.in., data=Accident_Training)
summary(lm_model1)
```
### Let's answer a couple of question:

* Is there a relationship between the predictor and the response? => Yes

* How strong is the relationship between the predictor and the response? => p
 -value is close to 0: relationship is strong

* Is the relationship between the predictor and the response positive or negative? => Positive

* In summary, the model suggests that there is a statistically significant positive relationship between Precipitation.in. and Distance.mi., meaning that an increase in precipitation is associated with an increase in distance of accident.

```{r}
# model 2
lm_model2 <- lm(Distance.mi. ~ Visibility.mi., data=Accident_Training)
summary(lm_model2)
```
* Is there a relationship between the predictor and the response? => Yes

* How strong is the relationship between the predictor and the response? => p
 -value is close to 0: relationship is strong

* Is the relationship between the predictor and the response positive or negative? => Negative

* In summary, this model suggests that visibility has a statistically significant effect on distance. This means that a decrease in visibility tends to lead to larger distance of the accident.


```{r}
# model 3
lm_model3 <- lm(Distance.mi. ~ Wind_Chill.F., data=Accident_Training)
summary(lm_model3)
```
* In summary, the model suggests a statistically significant relationship between the predictor (Wind_Chill) and the response variable (Distance)

### Let's summarize prediction data and calculate MAE (Mean Absolute Error) and MSE (Mean Squared Error) . MAE and MSE are both metrics commonly used to evaluate the performance of a regression model

```{r}
predictions1 <- predict(lm_model1, newdata = Accident_Testing)
predictions2 <- predict(lm_model2, newdata = Accident_Testing)
predictions3 <- predict(lm_model3, newdata = Accident_Testing)
summary(predictions1)
```

```{r}
summary(predictions2)
```
```{r}
summary(predictions3)
```
```{r}
MAE(y_pred = predictions1, y_true = Accident_Testing$Distance.mi.)
```

```{r}
MAE(y_pred = predictions2, y_true = Accident_Testing$Distance.mi.)
```
```{r}
MAE(y_pred = predictions3, y_true = Accident_Testing$Distance.mi.)
```
```{r}
MSE(y_pred = predictions1, y_true = Accident_Testing$Distance.mi.)
```

```{r}
MSE(y_pred = predictions2, y_true = Accident_Testing$Distance.mi.)
```
```{r}
MSE(y_pred = predictions3, y_true = Accident_Testing$Distance.mi.)
```


# Multiple Linear Regression


```{r}
multiplelm1 <- lm(Distance.mi.~., data=Accident_Training)
summary(multiplelm1)
```
```{r}
ypred <-predict(object = multiplelm1, newdata = Accident_Testing)
summary(ypred)
```


```{r}
MAE(y_pred = ypred, y_true = Accident_Testing$Distance.mi.)
```

```{r}
MSE(y_pred = ypred, y_true = Accident_Testing$Distance.mi.)
```

# Forward Stepwise regression

```{r}
intercept_only <- lm(Distance.mi. ~ 1, data= Accident_Training)
all <- lm(Distance.mi. ~., data=Accident_Training)
forward <- stepAIC (intercept_only, direction='forward',scope = formula(all))
```

```{r}
summary(forward)
```

>>In summary, the final iteration model includes predictors Pressure.in., Visibility.mi., Wind_Chill.F., Temperature.F., and Humidity..., and it has the lowest AIC among the considered models. This model is chosen as it strikes a balance between model complexity and goodness of fit. It can be noted that the improvement made by adding humidity had a very small improvement.

```{r}
ypred_forward <-predict(object = forward, newdata = Accident_Testing)
MAE(y_pred = ypred_forward, y_true = Accident_Testing$Distance.mi.)
```
```{r}
MSE(y_pred = ypred_forward, y_true = Accident_Testing$Distance.mi.)
```

# Backward Stepwise Regression

```{r}
backward <- stepAIC (all, direction='backward')
```
```{r}
summary(backward)
```

>> The output then shows the stepwise elimination of variables based on the AIC.The first step eliminates the variable "Precipitation.in.", the second step eliminates "Wind_Speed.mph."

```{r}
ypred_backward <-predict(object = backward, newdata = Accident_Testing)
MAE(y_pred = ypred_backward, y_true = Accident_Testing$Distance.mi.)
```

```{r}
MSE(y_pred = ypred_backward, y_true = Accident_Testing$Distance.mi.)
```


### What happens if we don't include Distance at all? Let's instead try to use Temperature as the response variable.

```{r}
multiplelm_withoutdistance <- lm(Temperature.F.~., data=Accident_Training[,2:8])
summary(multiplelm_withoutdistance)
```
```{r}
par(mfrow=c(2,2))
plot(multiplelm_withoutdistance)
```

